PCA on Finnish Municipalities

### 2022-04-11 MSA-Project.Rmd
Updated section on distribution plots

To analyze how variables are distributed, we use Kernel density estimation to estimate the probability densities for several variable groups.

Figure \@ref(fig:agekde) shows the probability density estimates for age groups. Age groups are shares of population aged below 15, 15 to 64, and over 64. We can see that ...

Figure \@ref(fig:housingkde) shows the probability density estimates for housing types. Housing types are shares of population living in either terraced or detached houses and in rental apartments. The Figure revels that ...

@@

```{r scoreplot1, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Score plot for all municipalities that have a top 5 absolute loading for any of PC1 and PC2", fig.align='left', optipng = '-o7'}

@@

ggplot(loads.reduced, aes(x = PC1, y = PC2, label = Area)) +
  geom_point() +
  geom_text_repel(size = 3) +
  xlim(c(-25,10)) +
  ylim(c(-20,15)) +
  ggtitle("PCA Scores, 1st and 2nd PC") +
  theme_light()

@@

```{r loadingsplot1, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Loading directions for PC1 and PC2", fig.align='left', optipng = '-o7'}

ggplot2::autoplot(muni.pca, 
                  alpha = 0,
                  loadings = T, 
                  loadings.colour = "black", 
                  loadings.label = T, 
                  loadings.label.colour = "red",
                  loadings.label.size = 2.5, 
                  loadings.label.repel = T) +
  scale_x_continuous(limits = c(-0.4, 0.4)) +
  scale_y_continuous(limits = c(-0.4, 0.4)) +
  labs(x = "Standardized PC1 (33.6% explained var.)",
       y = "Standardized PC1 (20.9% explained var.)",
       title = "PCA Loadings, 1st and 2nd PC") +
  theme_light()

@@

```{r scoreplot2, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Score plot for all municipalities that have a top 5 absolute loading for any of PC3 and PC4", fig.align='left', optipng = '-o7'}

@@

ggplot(loads.reduced, aes(x = PC3, y = PC4, label = Area)) +
  geom_point() +
  geom_text_repel(size = 3) +
  xlim(c(-5,5)) +
  ylim(c(-8,8)) +
  ggtitle("PCA Scores, 3rd and 4th PC") +
  theme_light()

@@

```{r loadingsplot2, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Loading directions for PC3 and PC4", fig.align='left', optipng = '-o7'}

ggplot2::autoplot(muni.pca,
                  x = 3,
                  y = 4,
                  alpha = 0,
                  loadings = T, 
                  loadings.colour = "black", 
                  loadings.label = T, 
                  loadings.label.colour = "red",
                  loadings.label.size = 2.5, 
                  loadings.label.repel = T) +
  scale_x_continuous(limits = c(-0.3, 0.3)) +
  scale_y_continuous(limits = c(-0.3, 0.3)) +
  labs(x = "Standardized PC3 (10.2% explained var.)",
       y = "Standardized PC4 (7.2% explained var.)",
       title = "PCA Loadings, 3rd and 4th PC") +
  theme_light()

>>>>

To analyze how variables are distributed, we use Kernel density estimation to estimate the probability densities for variables in some common groups.

Figure \@ref(fig:agekde) shows the probability density estimates for age groups. Age groups are shares of population aged below 15, 15 to 64, and over 64. We can see that Prop.Over64  has a wide distribution. The median is close to 30% but there are municipalities in both ends of the distribution. The distribution of Prop.Under15 has a lower variance around its mean value of ~15% but the distribution is clearly positively skewed as there are some municipalities with over 30% of below 15 year-old citizens (Liminka; 30.8%).

Figure \@ref(fig:housingkde) shows the probability density estimates for housing types. Housing types are shares of population living in either terraced or detached houses and in rental apartments. The Figure characterizes Finland well - the distribution of Prop.Households.Terr.Det show that there are lots of municipalities where ~9/10 citizens tend to live in terraced or detached houses. However, the distribution is highly negatively skewed so there are some municipalities where very few people live in those types of houses. On the other hand, Prop.Households.Rental is centered around ~20% but highly positively skewed. This means that while in average 1/5 of people live on rent, there are municipalities where more than half are living on rent. Overall, these results support to hypothesis that Finland is a country of many rural municipalities where people tend to live in terraced and detached houses, and a few big municipalities where the proportion of rental housing increases with city size and the degree of urbanization.

Similar distribution plots for employment measures and economic sectors are shown in the Appendix.

@@

```{r scoreplot1, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Score plot for all municipalities that have a top 5 absolute loading for any of PC1 and PC2", fig.align='left', optipng = '-o7', fig.dim=c(6,4)}

@@

ggplot(loads.reduced, aes(x = PC1, y = PC2, label = Area)) +
  geom_point() +
  geom_text_repel(size = 3) +
  xlim(c(-25,10)) +
  ylim(c(-20,15)) +
  ggtitle("PCA Scores") +
  theme_light()

@@

```{r loadingsplot1, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Loading directions for PC1 and PC2", fig.align='left', optipng = '-o7', fig.dim=c(6,4)}

ggplot2::autoplot(muni.pca, 
                  alpha = 0,
                  loadings = T, 
                  loadings.colour = "black", 
                  loadings.label = T, 
                  loadings.label.colour = "red",
                  loadings.label.size = 2.5, 
                  loadings.label.repel = T) +
  scale_x_continuous(limits = c(-0.4, 0.4)) +
  scale_y_continuous(limits = c(-0.4, 0.4)) +
  labs(x = "Standardized PC1 (33.6% explained var.)",
       y = "Standardized PC1 (20.9% explained var.)",
       title = "PCA Loading Directions") +
  theme_light()

@@

```{r scoreplot2, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Score plot for all municipalities that have a top 5 absolute loading for any of PC3 and PC4", fig.align='left', optipng = '-o7', fig.dim=c(6,4)}

@@

ggplot(loads.reduced, aes(x = PC3, y = PC4, label = Area)) +
  geom_point() +
  geom_text_repel(size = 3) +
  xlim(c(-4,5)) +
  ylim(c(-8,8)) +
  ggtitle("PCA Scores") +
  theme_light()

@@

```{r loadingsplot2, warning=FALSE, echo=FALSE, message=FALSE, fig.cap="Loading directions for PC3 and PC4", fig.align='left', optipng = '-o7', fig.dim=c(6,4)}

ggplot2::autoplot(muni.pca,
                  x = 3,
                  y = 4,
                  alpha = 0,
                  loadings = T, 
                  loadings.colour = "black", 
                  loadings.label = T, 
                  loadings.label.colour = "red",
                  loadings.label.size = 2.5, 
                  loadings.label.repel = T) +
  scale_x_continuous(limits = c(-0.3, 0.3)) +
  scale_y_continuous(limits = c(-0.3, 0.3)) +
  labs(x = "Standardized PC3 (10.2% explained var.)",
       y = "Standardized PC4 (7.2% explained var.)",
       title = "PCA Loading Directions") +
  theme_light()
